### 2025-05-28 paper/relatedwork.tex
Address some reviewer A comments.

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
(ii) if the
language does not perform as expected, what are the
options to solve this?  Unfortunately, in many cases the answer to
both questions is unsatisfying.  Most languages do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Note that while our work
does not give a formal proof that AD implementation is correct (it is
not clear how to formulate correcntess criteria for it), the entire
implementatuion of AD is about 20 lines of code, so it is easy to
understand and modify.

@@

Thiemann and Chakravarty demonstrated a prototype embedding of
Accelerate~\cite{10.1145/1926354.1926358} (a Haskell library for
accelerator programming) in Agda~\cite{thiemann2013agda}. Their
approach was based on dynamic code generation by invoking Accelerate
through Agda's foreign function interface, which posed various
unresolved implementation challenges. Even though our for support rank
polymorphism results in having even more complicated types, our code
extraction, which is essentially ahead-of-time compilation, is fairly
unproblematic.

@@

Verifying index operations is a classic application of dependent
types, as seen in for example Dependent
ML~\cite{10.1145/292540.292560}, although most work assumes static
ranks. Our target language, Futhark, supports size-dependent
types~\cite{10.1145/3609024.3609412}, but this information is used
only to ensure shape conformance (e.g., that the operands to a matrix
multiplication have the same size), not to ensure the correctness of
array indexing. Single-Assignment C~\cite{sac2} (SaC) supports rank
polymorphism, including dynamic ranks, but does not guarantee that
absence of indexing errors.

Gibbons showed how to express rank polymorphism in Haskell, through
the use of \emph{Naperian functors}~\cite{10.1145/2976022.2976023}.
This form of rank polymorphism is more limited than ours; in
particular, it imposes static ranks. Further, Gibbons' work focused
solely on the embedding in Haskell, while our system also supports
code extraction and demonstrates good real-world performance.
Similarly, Remora~\cite{rank-poly} is a dependently typed
rank-polymorphic language that also supports only static ranks.

>>>>

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
(ii) if the
language does not perform as expected, what are the
options to solve this?  Unfortunately, in many cases the answer to
both questions is unsatisfying.  Most languages do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Note that while our work
does not give a formal proof that AD implementation is correct (it is
not clear how to formulate correctness criteria for it), the entire
implementation of AD is about 20 lines of code, so it is easy to
understand and modify.

@@

Thiemann and Chakravarty demonstrated a prototype embedding of
Accelerate~\cite{10.1145/1926354.1926358} (a Haskell library for
accelerator programming) in Agda~\cite{thiemann2013agda}. Their
approach was based on dynamic code generation by invoking Accelerate
through Agda's foreign function interface, which posed various
unresolved implementation challenges. Even though our support for rank
polymorphism results in having even more complicated types, our code
extraction, which is essentially ahead-of-time compilation, is fairly
unproblematic.

@@

Verifying index operations is a classic application of dependent types, as seen
in for example Dependent ML~\cite{10.1145/292540.292560}, although most work
assumes static ranks. Our target language, Futhark, supports size-dependent
types~\cite{10.1145/3609024.3609412}, but this information is used only to
ensure shape conformance (e.g., that the operands to a matrix multiplication
have the same size), and does not guarantee that array indexing is in-bounds.
Single-Assignment C~\cite{sac2} (SaC) supports rank polymorphism, including
dynamic ranks, but does not guarantee that absence of indexing errors.

Gibbons showed how to express rank polymorphism in Haskell, through the use of
\emph{Naperian functors}~\cite{10.1145/2976022.2976023}. This form of rank
polymorphism is more limited than ours; in particular, it imposes static ranks.
This is also the case for libraries that provide rank polymorphic array
operations, such as Accelerate~\cite{10.1145/1926354.1926358} and
Repa~\cite{10.1145/1932681.1863582}. Further, Gibbons' work focused solely on
the embedding in Haskell, while our system also supports code extraction and
demonstrates good real-world performance. Similarly, Remora~\cite{rank-poly} is
a dependently typed rank-polymorphic language that also supports only static
ranks.
